preciseTAD: A machine learning framework for precise 3D domain boundary prediction at base-level resolution

### 2020-09-17 Manuscript/manuscript.Rmd
added cross cell line prediction last paragraph manuscript; editted figures

The majority of boundaries predicted by the Arrowhead-trained *preciseTAD* model represented a subset of the larger group of boundaries predicted by the Peakachu-trained model. We found that 88.8% and 95.8% of our predicted TAD boundaries were overlapped by predicted loop boundaries for GM12878 and K562, respectively (Supplementary Figure 9). This is expected as loop boundaries (Peakachu) are more abundant, while TAD boundaries (Arrowhead) represent the narrower level of the 3D chromatin organization. Together with the results in Figure 4, the high overlap between boundaries predicted by Arrowhead- and Peakachu-trained models suggests that TAD and loop boundaries may be driven by similar molecular mechanisms.

@@

*preciseTAD* offers flexibility in controlling both the number of predicted boundaries and the distance between them. The two primary parameters are the probability threshold $t$ and $\epsilon$ (referred to as eps; parameter of DBSCAN). The combination of these two quantities changes the resulting number of predicted boundaries from *preciseTAD*. Lower values of $t$ and $\epsilon$ will result in more clusters of bases, and therefore, more boundaries. As a heuristic, we evaluated the pairwise combination of 3 different thresholds ($t=(0.975,0.99,1.0)$) and 6 different $\epsilon$ values ($\epsilon=(1000,5000,10000,15000,20000,25000)$)). We found that the normalized overlaps - calculated as the total number of ChIP-seq peaks that overlapped within a given flanked boundary, divided by the number of boundaries - between top TFBS sites and flanked *preciseTAD* boundaries converged for combinations of $t=1.0$ and $\epsilon=10000$ (Supplementary Figure 10).

@@

**Supplementary Figure 1.** Density curves of distances before (red) and after (blue) performing a $log_{2}$ transformation across 10 kb, 25 kb, 50 kb, and 100 kb data resolutions for both the (A) GM12878 and (B) K562 cell lines, respectively. Distances are measured as the number of bases from the center of a genomic bin to the nearest genomic annotation center. Each density curve represents an individual genomic annotation (77 total).

**Supplementary Figure 2** Psuedocode of the *preciseTAD* algorithm.

@@

**Supplementary Figure 3** (A) Averaged balanced accuracies are compared across resolution, within each predictor-type: Signal, OC, OP, and Distance, and across resampling techniques: no resampling (None; red), random over-sampling (ROS; green), random under-sampling (RUS; blue), and synthetic minority over-sampling (SMOTE; purple) when using Arrowhead ground truth boundaries for K562. (B) Averaged balanced accuracies are compared for GM12878 and (C) K562  within each predictor-type: Signal, OC, OP, and Distance, and across resampling technique: no resampling (None; red), random over-sampling (ROS; green), random under-sampling (RUS; blue), and synthetic minority over-sampling (SMOTE; purple). Error bars indicate standard deviation away from the mean performance across each holdout chromosome used for testing.

**Supplementary Figure 4.** (A) Barplots comparing performances of TAD (Arrowhead) and loop (Peakachu) boundary prediction models using K562-specific histone modifications (HM), chromatin states (BroadHMM), transcription factor binding sites (TFBS), in addition to a model containing all three classes (ALL). (B) Recursive feature elimination (RFE) analysis used to select the optimal number of predictors. Error bars represent 1 standard deviation from the cross-validated accuracy across each holdout chromosome. (C) Clustered heatmap of the predictive importance for the union of the top 8 most predictive chromosome-specific TFBSs. The columns represent the holdout chromosome excluded from the training data. (D) Receiver operating characteristic (ROC) curves and the corresponding average area under the curves (AUCs) when training and testing on the same cell line (blue. Arrowhead ground truth; red, Peakachu ground truth) versus across cell lines (black, dashed). The curves represent the average sensitivities and specificities across each holdout chromosome. The shaded areas around each curve represent $\pm$ 1 standard deviation aways from the average.

@@

**Supplementary Figure 6.** *preciseTAD* boundaries are more enriched for known molecular drivers of 3D chromatin. Enriched heatmaps comparing the enrichment of Arrowhead called TAD boundaries vs *preciseTAD*-predicted TAD boundaries around key transcription factor binding sites known to contribute to the 3D chromatin architecture for (A) GM12878 and (B) K562.

**Supplementary Figure 7.** *preciseTAD* boundaries are more enriched for known molecular drivers of 3D chromatin. Enriched heatmaps comparing the enrichment of Peakachu chromatin loop boundaries vs *preciseTAD*-predicted chromatin boundaries around key transcription factor binding sites known to contribute to the 3D chromatin architecture for (A) GM12878 and (B) K562.

**Supplementary Figure 8.** *preciseTAD* boundaries are spatially closer to known molecular drivers of 3D chromatin. Boxplots comparing the log2 genomic distances from predicted and called boundaries to the nearest (A) GM12878-specific and (B) K562-specific CTCF, RAD21, SMC3, and ZNF143 transcription factor binding sites. The p-value is derived from the Wilcoxon Rank Sum test.

**Supplementary Figure 9.** *preciseTAD*-predicted TAD boundaries using Arrowhead are nested within *preciseTAD*-predicted chromatin loop boundaries using Peakachu. Venn diagrams illustrating the agreement between boundaries predicted using Arrowhead- and Peakachu-trained models for (A) GM12878 and (B) K562. Domain boundaries were flanked by resolution (5 kb for models using Arrowhead and 10 kb for Peakachu).

**Supplementary Figure 10.** Linecharts illustrating the normalized enrichment (NE) between known molecular drivers of 3D chromatin and resolution-flanked *preciseTAD*-predicted boundaries for different combinations of thresholds (t) and epsilon parameter values (eps). Data from GM12878 and K562 cell lines, chromosome 22, at 5kb resolution was used. NE was calculated as the total number of ChIP-seq peaks that overlapped within a given flanked boundary, divided by the number of boundaries that were predicted, and averaged over the number of annotations included in the model. Error bars indicate standard deviation away from the mean.

>>>>

The majority of boundaries predicted by the Arrowhead-trained *preciseTAD* model represented a subset of the larger group of boundaries predicted by the Peakachu-trained model. We found that 88.8% and 95.8% of our predicted TAD boundaries were overlapped by predicted loop boundaries for GM12878 and K562, respectively (Supplementary Figure 9). This is expected as loop boundaries (Peakachu) are more abundant, while TAD boundaries (Arrowhead) represent the narrower level of the 3D chromatin organization. Together with the results in Figure 4, the high overlap between boundaries predicted by Arrowhead- and Peakachu-trained models suggests that TAD and loop boundaries may be driven by similar molecular mechanisms. Guided by the high predictive performance when training and testing on different cell lines (Figure 4D), we opted to evaluate whether *preciseTAD* could be leveraged to predict boundaries using unseen base level cell line-specific annotation data. To do so, we evaluated two separate scenarios for each type of ground truth boundary: 1) training on GM12878 and predicting boundaries on GM12878 (GM on GM) vs training on K562 and predicting on GM12878 (K on GM), and 2) training on K562 and predicting boundaries on K562 (K on K) vs training on GM12878 and predicting boundaries on K562 (GM on K). Using Arrowhead ground truth, 76% (J=0.701) and 81% (J=0.751) of predicted boundaries were overlapped in each scenario, respectively (Supplementary Figure 10A, 10B). Likewise, when using Peakachu ground truth, there was 85% (J=0.705) and 88% (J=0.759) of overlap between predicted boundaries, respectively (Supplementary 10C, 10D). Furthermore, boundaries predicted on unseen annotation data exhibited similar levels of enrichment for CTCF, RAD21, SMC3, and ZNF143 as did those trained and predicted on the same cell line. These results open up the possibility of using preciseTAD to predict TAD or loop boundaries for cell lines that do not currently have Hi-C data.

@@

*preciseTAD* offers flexibility in controlling both the number of predicted boundaries and the distance between them. The two primary parameters are the probability threshold $t$ and $\epsilon$ (referred to as eps; parameter of DBSCAN). The combination of these two quantities changes the resulting number of predicted boundaries from *preciseTAD*. Lower values of $t$ and $\epsilon$ will result in more clusters of bases, and therefore, more boundaries. As a heuristic, we evaluated the pairwise combination of 3 different thresholds ($t=(0.975,0.99,1.0)$) and 6 different $\epsilon$ values ($\epsilon=(1000,5000,10000,15000,20000,25000)$)). We found that the normalized overlaps - calculated as the total number of ChIP-seq peaks that overlapped within a given flanked boundary, divided by the number of boundaries - between top TFBS sites and flanked *preciseTAD* boundaries converged for combinations of $t=1.0$ and $\epsilon=10000$ (Supplementary Figure 11).

@@

**Supplementary Figure 1. $log_{2}$ transformation normalizes genomic distance.** Density curves of distances before (red) and after (blue) performing a $log_{2}$ transformation across 10 kb, 25 kb, 50 kb, and 100 kb data resolutions for both the (A) GM12878 and (B) K562 cell lines, respectively. Distances are measured as the number of bases from the center of a genomic bin to the nearest genomic annotation center. Each density curve represents an individual genomic annotation (77 total).

**Supplementary Figure 2.** Psuedocode of the *preciseTAD* algorithm.

@@

**Supplementary Figure 3. Distance-type predictors combined with random under-sampling at 5 kb resolution genomic bins yielded optimal boundary region predictive performance.** (A) Averaged balanced accuracies are compared across resolution, within each predictor-type: Signal, OC, OP, and Distance, and across resampling techniques: no resampling (None; red), random over-sampling (ROS; green), random under-sampling (RUS; blue), and synthetic minority over-sampling (SMOTE; purple) when using Arrowhead ground truth boundaries for K562. (B) Averaged balanced accuracies are compared for GM12878 and (C) K562  within each predictor-type: Signal, OC, OP, and Distance, and across resampling technique: no resampling (None; red), random over-sampling (ROS; green), random under-sampling (RUS; blue), and synthetic minority over-sampling (SMOTE; purple). Error bars indicate standard deviation away from the mean performance across each holdout chromosome used for testing.

**Supplementary Figure 4. Binding of SMC3, RAD21, CTCF, and ZNF143 transcription factors accurately predicts TAD and loop boundaries in K562 cell line.** (A) Barplots comparing performances of TAD (Arrowhead) and loop (Peakachu) boundary prediction models using K562-specific histone modifications (HM), chromatin states (BroadHMM), transcription factor binding sites (TFBS), in addition to a model containing all three classes (ALL). (B) Recursive feature elimination (RFE) analysis used to select the optimal number of predictors. Error bars represent 1 standard deviation from the cross-validated accuracy across each holdout chromosome. (C) Clustered heatmap of the predictive importance for the union of the top 8 most predictive chromosome-specific TFBSs. The columns represent the holdout chromosome excluded from the training data. (D) Receiver operating characteristic (ROC) curves and the corresponding average area under the curves (AUCs) when training and testing on the same cell line (blue. Arrowhead ground truth; red, Peakachu ground truth) versus across cell lines (black, dashed). The curves represent the average sensitivities and specificities across each holdout chromosome. The shaded areas around each curve represent $\pm$ 1 standard deviation aways from the average.

@@

**Supplementary Figure 6. *preciseTAD* boundaries are more enriched for known molecular drivers of 3D chromatin.** Enriched heatmaps comparing the enrichment of Arrowhead called TAD boundaries vs *preciseTAD*-predicted TAD boundaries around key transcription factor binding sites known to contribute to the 3D chromatin architecture for (A) GM12878 and (B) K562.

**Supplementary Figure 7. *preciseTAD* boundaries are more enriched for known molecular drivers of 3D chromatin.** Enriched heatmaps comparing the enrichment of Peakachu chromatin loop boundaries vs *preciseTAD*-predicted chromatin boundaries around key transcription factor binding sites known to contribute to the 3D chromatin architecture for (A) GM12878 and (B) K562.

**Supplementary Figure 8. *preciseTAD* boundaries are spatially closer to known molecular drivers of 3D chromatin.** Boxplots comparing the log2 genomic distances from predicted and called boundaries to the nearest (A) GM12878-specific and (B) K562-specific CTCF, RAD21, SMC3, and ZNF143 transcription factor binding sites. The p-value is derived from the Wilcoxon Rank Sum test.

**Supplementary Figure 9.*preciseTAD*-predicted TAD boundaries are nested within *preciseTAD*-predicted chromatin loop boundaries.** Venn diagrams illustrating the agreement between boundaries predicted using Arrowhead- and Peakachu-trained models for (A) GM12878 and (B) K562. Domain boundaries were flanked by resolution (5 kb for models using Arrowhead and 10 kb for Peakachu).

**Supplementary Figure 10. *preciseTAD* accurately predicts boundaries on cell lines using unseen annotation data.** Venn diagrams comparing flanked predicted boundaries using Arrowhead ground truth to: (A) train on GM12878 and predict on GM12878 (GM on GM) vs train on K562 and predict on GM12878 (K on GM), (B) train on K562 predict on K562 (K on K) vs train on GM12878 predict on K562 (GM on K). (C-D) Using Peakachu ground truth. The corresponding profilePlots (deepTools version 2.0) are provided to the right. Predicted boundaries using Arrowhead ground truth were flanked by 5 kb, while boundaries predicted using Peakachu ground truth were flanked by 10 kb.

**Supplementary Figure 11. Normalized Enrichment levels converge for combinations on t=1.0 and $\epsilon=10000$.** Linecharts illustrating the normalized enrichment (NE) between known molecular drivers of 3D chromatin and resolution-flanked *preciseTAD*-predicted boundaries for different combinations of thresholds (t) and epsilon parameter values (eps) using data from GM12878 (A) and K562 (B) cell lines, chromosome 22, at 5kb resolution was used. NE was calculated as the total number of ChIP-seq peaks that overlapped within a given flanked boundary, divided by the number of boundaries that were predicted, and averaged over the number of annotations included in the model. Error bars indicate standard deviation away from the mean.
